Abstraction and Invariance for Algebraically Indexed Types

### 2012-07-09 paper/section-instantiations.tex
Small fixes in intro; fixed error in AG axioms of vector addition; new type isomorphisms maps

\section{Geometric Consequences of Abstraction}
\label{sec:instantiations}

\fixme{explain the general setup, with a set of operations for each
  instantiation.}

\begin{lemma}
  The thing about $\Gamma_\Geom$.
\end{lemma}

@@

Consider the type of the area of a triangle function we defined in
\exref{ex:area-of-triangle-1}:
\begin{displaymath}
  \begin{array}{@{}l}
    \mathrm{area} : \forall B\mathord:\SynGL{2}, t\mathord:\SynTransl{2}.\ \\
    \hspace{0.8cm}\tyPrim{vec}{B, t} \to \tyPrim{vec}{B, t} \to \tyPrim{vec}{B, t} \to \tyPrim{real}{|\det B|}
  \end{array}
\end{displaymath}
By \thmref{thm:abstraction}, we can derive the following free theorem. For all $B \in \GL{2}$, $\vec{t} \in \Transl{2}$, and $\vec{x}, \vec{y}, \vec{z} \in \mathbb{R}^2$, we have 
\begin{displaymath}
  |\det B|(\tmSem{\mathrm{area}}\ \vec{x}\ \vec{y}\ \vec{z}) = \tmSem{\mathrm{area}}\ (B\vec{x} + \vec{t})\ (B\vec{y} + \vec{t})\ (B\vec{z} + \vec{t})
\end{displaymath}
Thus, directly from the type of the $\mathrm{area}$ function, we can
see that its index-erasure semantics is (a) invariant under
translations, and (b) if the inputs are subjected to a linear
transformation $B$, the output varies with the absolute value of the
determinant of $B$.

@@

\subsection{Isomorphisms for Types Indexed by Abelian Groups}
\label{sec:types-indexed-abelian-groups}

\begin{enumerate}
\item Two cases: when we have a total ``division'' operation, and when
  we don't.
\item Type isomorphisms, based on Smith Normal Form
% \item Non-definability results, based on the non-compositional
%   relational environments
% \item Alteration of the non-definability results in the light of
%   addition of a square root operation
% \item Mention the variation we can obtain by changing some types,
%   e.g. the type of $\mathrm{abs}$. Also, the complex numbers example.
\end{enumerate}

>>>>

\section{Geometric Consequences of Abstraction}
\label{sec:instantiations}
We now instantiate our general framework %of the previous section 
with
the indexing theory of \autoref{sec:motivating-examples}, and present
more general and formally-justified theorems for free, type
isomorphisms, and non-definability results.

%\begin{example}[Two-dimensional Geometry]

We define the ``standard'' model of the indexing theory for the
two-dimensional geometry example as follows. Each of the sorts is
interpreted just as its semantic counterpart: $\semSort{\SynTransl{2}}
= \Transl{2}$, $ \semSort{\SynGL{2}} = \GL{2}$, $
\semSort{\SynOrth{2}} = \Orth{2}$, and $ \semSort{\SynGL{1}} =
\GL{1}$.
%  \begin{displaymath}
%    \begin{array}{@{}llll}
%      \semSort{\SynTransl{2}} = \Transl{2}
%      &
%      \semSort{\SynGL{2}} = \GL{2}
%      &
%      \semSort{\SynOrth{2}} = \Orth{2}
%      &
%      \semSort{\SynGL{1}} = \GL{1}
%    \end{array}
%  \end{displaymath}
  Each of the index operations (e.g.,~the group structure and
  determinant) is interpreted by the intended semantic operation, and
  clearly satisfy the axioms in \exref{ex:two-dim-geo-axioms}. The
  assignment of relations for the primitive types $\tyPrimNm{vec}$ and
  $\tyPrimNm{real}$ is given by: % the following definitions:
  \begin{displaymath}
    \begin{array}{ll}
      R^{\bullet}_{\tyPrimNm{vec}} = \{( \vec{0}, \vec{0} ) \}&
      R_{\tyPrimNm{vec}}(B,\vec{t}) = \{ (\vec{v}, B\vec{v} + \vec{t}) \sepbar \vec{v} \in \mathbb{R}^2 \} \\
      R^{\bullet}_{\tyPrimNm{real}} = \{(0,0)\} &
      R_{\tyPrimNm{real}}(k) = \{ (x, kx) \sepbar x \in \mathbb{R} \}
    \end{array}
  \end{displaymath}
  The definitions of $R_{\tyPrimNm{vec}}$ and $R_{\tyPrimNm{real}}$
  %are as expected. % given the informal definitions we gave in
formalise the informal ones from~\autoref{sec:motivating-examples}, and 
  the definitions of %We have set
  $R^\bullet_{\tyPrimNm{vec}}$ and $R^\bullet_{\tyPrimNm{real}}$ 
  %to relate $\vec{0}$ and $0$ to themselves to 
  account for the
  polymorphic zero constants we have assumed.
%\end{example}

%\begin{lemma}
%  The thing about $\Gamma_\Geom$.
%\end{lemma}

@@

Consider the type of the area %of a triangle 
function from
\exref{ex:area-of-triangle-1}:
\begin{displaymath}
  \begin{array}{@{}l}
    \mathrm{area} : \forall B\mathord:\SynGL{2}, t\mathord:\SynTransl{2}.\ \\
    \hspace{0.8cm}\tyPrim{vec}{B, t} \to \tyPrim{vec}{B, t} \to \tyPrim{vec}{B, t} \to \tyPrim{real}{|\det B|}
  \end{array}
\end{displaymath}
By \thmref{thm:abstraction}, we can derive the following free theorem. For all $B \in \GL{2}$, $\vec{t} \in \Transl{2}$, and $\vec{x}, \vec{y}, \vec{z} \in \mathbb{R}^2$, we have 
\begin{displaymath}
  |\det B|(\tmSem{\mathrm{area}}\ \vec{x}\ \vec{y}\ \vec{z}) = \tmSem{\mathrm{area}}\ (B\vec{x} + \vec{t})\ (B\vec{y} + \vec{t})\ (B\vec{z} + \vec{t})
\end{displaymath}
Thus, directly from the type of the $\mathrm{area}$ function, we can
see that its index-erasure semantics is (a) invariant under
translations, and (b) if the inputs are subjected to a linear
transformation $B$, the output varies with the absolute value of the
determinant of $B$.

@@

\subsection{Type Isomorphisms}
\label{sec:types-indexed-abelian-groups}
We say that two types $A$ and $B$ are isomorphic, and write $A \cong
B$, if there exist maps between them that are mutually inverse with
respect to contextual equivalence, i.e., if there are terms $\vdash I
: A \to B$ and $\vdash J : B\to A$ such that $x:A\vdash
J(I(x))\stackrel{ctx}\approx x : A$ and $y:B\vdash
I(J(y))\stackrel{ctx}\approx y : B$.
%\[
%x:A\vdash J(I(x))\stackrel{ctx}\approx x : A
%\text{ and }y:B\vdash I(J(y))\stackrel{ctx}\approx y : B.
%\]
That $\cong$ is a congruence relation on types is straightforward. We
can also derive isomorphisms that are independent of the indexing
theory, such as $A\times B \cong B \times A$, and $\forall
i\mathord:s.A\to B\cong A\to\forall i\mathord:s.B$ for $i$ not free in
$A$.

Types indexed by abelian groups induce a particularly rich theory of
type isomorphisms; previous work on
units-of-measure~\cite{kennedy97relational} relates these to
Buckingham's theorem from dimensional analysis. Here we consider the
additive abelian groups of translations and the multiplicative abelian
group of scalings.

\paragraph{Translations} 
Consider first the group $\Transl{2}$ of translations from
\autoref{sec:type-system-geom-intro} and 
\autoref{sec:affine-vector-ops}. For any quantifier-free type $A$
we have: % the isomorphism:
\begin{eqnarray*}
\forall t\mathord:\SynTransl{2}.\tyPrim{vec}{t}\to A
& \cong &
A[0/t]
\end{eqnarray*}
where $A[0/t]$ denotes substitution of the identity translation $0$
for index variable $t$ in type $A$. This isomorphism formalises the
equivalence between \emph{coordinate-free} geometry, expressed by a
polymorphic type whose first argument can be thought of as the
\emph{origin} with respect to which $A$ is specified, and a
\emph{coordinate-based} geometry, expressed by the type $A[0/t]$, in
which the origin is fixed at $(0,0)$. It is sometimes said that an
affine space is a vector space that's forgotten its origin; we have
captured this in a type isomorphism.

\exref{ex:type-iso} is a special case of this isomorphism; another instance is the
type of vector addition:
\begin{eqnarray*}
\forall t_1\mathord:\SynTransl{2}.\forall t_2\mathord:\SynTransl{2}.\tyPrim{vec}{t_1}\to\tyPrim{vec}{t_2}\to\tyPrim{vec}{t_1+t_2}\\
\cong\quad
\forall t_1\mathord:\SynTransl{2}.\tyPrim{vec}{t_1}\to\tyPrim{vec}{t_1}\quad\cong\quad
\tyPrim{vec}{0}
\end{eqnarray*}
\noindent
\newcommand{\transup}[1]{\uparrow_{#1}}
\newcommand{\transdn}[1]{\downarrow_{#1}}
The maps associated with this isomorphism are 
$I = \lambda x. x\ [0]\ 0$ and $J = \lambda y. \Lambda t.\lambda v. \transdn A (y)$,
%\begin{eqnarray*}
%I = \lambda x. x\ [0]\ 0 & \text{ and } &
%J = \lambda y. \Lambda t.\ \lambda v. \trans A v t(y)
%\end{eqnarray*}
where terms $\transdn A : A\to A[0/t]$ and $\transup A : A[0/t] \to A$ are defined by induction over the syntax of $A$:
\begin{eqnarray*}
\transdn{1} & = & \lambda x. x\\ 
\transdn{A_1\times A_2} & = & \lambda p. (\transdn{A_1}(\pi_1 p),\transdn{A_2}(\pi_2 p))\\ 
\transdn{A_1 + A_2} & = & \lambda x.
\textrm{case}\ x\ \textrm{of}\ \textrm{inl}\ y.\textrm{inl}(\transdn{A_1}y)); \textrm{inr}\ y.\textrm{inr}(\transdn{A_2}y)\\ 
\transdn{A_1\to A_2} & = & \lambda f. \lambda x. (\transdn{A_2} (f(\transup{A_1}(x))))\\ 
\transdn{\tyPrim{vec}{e}} & = & \lambda x. x - kv\\ 
\transup{\tyPrim{vec}{e}} & = & \lambda x. x + kv\\ & & \text{ if }e \equiv k t + e'\text{ for
  some }k\in\setOfIntegers \text{ and }e'\text{ with }t\notin e'\\
\end{eqnarray*}
To show $I(J(y))$ is contextually equivalent to $y$ is an easy
induction on $A$ using the fact that the vector $0$
is an identity for vector addition. To show $J(I(x))$ is contextually
equivalent to $x$ we must apply the abstraction theorem and use
the fact that $\tmSem{\transup A}$ and $\tmSem{\transdn A}$ 
considered as relations are simply instances of $\rsem A{}{}$.
%\[
%$\rsem A {t \mapsto v} {} = \{ (x, \tmSem{\transup A (x)}) \mid x \in
%\tySem A\}$.
%\]

\paragraph{Scalings}
Now consider the group $\GL{1}$ of scalings. This time, we can treat
the real-valued argument to a function as a `unit of measure' with
which to scale the result. 
%The small fly in the ointment is that the
Although the argument might be zero, and this cannot be used for scaling,
%Nevertheless, 
we still can obtain the following slightly more complicated
isomorphism:
\begin{eqnarray*}
\forall s\mathord:\SynGL{1}.\tyPrim{real}{s}\to A
& \cong &
A[1/s]\times \forall s\mathord:\SynGL{1}.A
\end{eqnarray*}
The type of $\mathrm{areaCircle}$ from the introduction is one instance:
\begin{eqnarray*}
\forall s\mathord:\SynGL{1}.\tyPrim{real}{s}\to\tyPrim{real}{s\cdot
  s}\\ \quad\cong\quad \tyPrim{real}{1}\times(\forall
s\mathord:\SynGL{1}.\tyPrim{real}{s\cdot s})\\ \quad\cong\quad
\tyPrim{real}{1}\times 1\quad\cong\quad \tyPrim{real}{1}
\end{eqnarray*}
%where %the isomorphism 
The abstraction theorem is used to prove
$\forall s\mathord:\SynGL{1}.\tyPrim{real}{s\cdot s}\cong 1$.
% is proved using
%the abstraction theorem.

%\begin{enumerate}
%\item Two cases: when we have a total ``division'' operation, and when
%  we don't.
%\item Type isomorphisms, based on Smith Normal Form
% \item Non-definability results, based on the non-compositional
%   relational environments
% \item Alteration of the non-definability results in the light of
%   addition of a square root operation
% \item Mention the variation we can obtain by changing some types,
%   e.g. the type of $\mathrm{abs}$. Also, the complex numbers example.
%\end{enumerate}

### 2012-07-09 paper/section-introduction.tex
Small fixes in intro; fixed error in AG axioms of vector addition; new type isomorphisms maps

In more complex examples the invariants \emph{compose} in various
ways, with the compositions satisfy algebraic laws. For example, 
we can assign a function that computes the area of a circle given its
radius the following polymorphic type:
\[
\mathrm{areaCircle} : \forall\alpha\mathord:\SynScal.\tyPrim{real}{\alpha}\to
\tyPrim{real}{\alpha\cdot\alpha}
\]
This captures the fact that the area of a circle varies as the square
of its radius, i.e., for any $s>0$, $\mathrm{areaCircle}\;(s\cdot r) =
s^2\cdot \mathrm{areaCircle}\;r$.  Here, $\alpha$ is interpreted as
the \emph{units of measure} of the argument to $\mathrm{areaCircle}$,
and `$\cdot$' composes units using the product. We can also add an
inverse operation and identity unit of measure, and then impose the
algebraic laws of abelian groups. This permits identification of, for
example, $\tyPrim{real}{\alpha\cdot\alpha^{-1}}$ with the type
$\tyPrim{real}{1}$ of dimensionless constants.

\paragraph{Abstraction}
In his original paper on parametricity, Reynolds asserted that
\emph{type structure is a syntactic discipline for enforcing levels of
  abstraction}.  We see something analogous here: if all primitive
operations are given types that reflect their behaviour under
translation, then there is no way to `break' this property. For
example, there is no way that $\mathrm{areaTri}$ can depend on the
actual coordinates of its inputs. Furthermore, the distinction between
points and vectors that is often enforced through abstract data
types~\cite{CGAL} is captured here by indices instead. For example,
the operation that takes two points and computes their vector
`difference' can be assigned the type
$\forall\alpha\mathord:\SynTransl.\tyPrim{vec}\alpha\times\tyPrim{vec}\alpha\to\tyPrim{vec}0$,
reflecting the invariance of the result (a pure vector) under
translations of the point arguments. As a result through types alone
we can, in essence, derive so-called \emph{coordinate-free}
geometry~\cite{CFGADT}.

@@

\paragraph{Relational parametricity}
To derive such invariance and abstraction properties of types, we
adopt the techniques of relational parametricity. Over an underlying
index-erasure semantics we construct binary relations parameterised by
an environment $\rho$ that describes how values of primitive type are
related according to their indices.  For example, values $v$ and $w$
of type $\tyPrim{real}\alpha$ are related when $v$ ``scales to'' $w$
according to an interpretation of $\alpha$ (i.e., $w=\rho(\alpha)\cdot
v$).  Values of polymorphic type are related exactly when they are
related for all possible interpretations of the quantified
variable. For example, values $v$ and $w$ of type
$\forall\alpha\mathord:\SynTransl.\tyPrim{vec}\alpha\to
\tyPrim{vec}\alpha$ are related when they are related at type
$\tyPrim{vec}\alpha\to\tyPrim{vec}\alpha$ for all translations
$t\in\Transl{2}$ associated with $\alpha$.

@@

As it happens, the above examples can all be expressed using unary
functions rather than binary relations: we can write functions that
scale, or translate, one value to another according to their type,
even for higher-order function types. Moreover, the interpretation of
indices can be built up compositionally, by recursion over the
syntactic structure. For example, we have $\rho(\alpha\cdot\alpha) =
\rho(\alpha)\cdot\rho(\alpha)$ where product of units-of-measure is
interpreted by product of scale factors. For more sophisticated
applications, however, we need the fully generality of binary
relations. First, as is also the case with type polymorphism, it
provides the appropriate meaning for nested quantifiers in types such
as
$(\forall\alpha\mathord:\SynScal.\tyPrim{real}\alpha\to\tyPrim{real}\alpha)\to\tyBool$.
More interestingly, particular applications sometimes require
relations that are not built compositionally---we can, for instance,
use such relations to prove that there are no non-trivial programs of
type
$\forall\alpha\mathord:\SynScal.\tyPrim{real}{\alpha\cdot\alpha}\to\tyPrim{real}\alpha$.
Finally, there are interesting applications in which even the basic
relations are not functional. In a type system in which the index in
$\tyPrim{real}\alpha$ is interpreted not as a unit of measure, but as
a measure of \emph{closeness}, two values $x$ and $y$ of this type are
related if $|x-y| < \rho(\alpha)$ for a positive real number
$\rho(\alpha)$.  Rather beautifully, the standard notion of uniform
continuity can then be expressed as a type:
\begin{displaymath}
  \forall \epsilon \mathord: \mathsf{R}^{>0}.\ \exists \delta\mathord: \mathsf{R}^{>0}.\ \tyPrim{real}{\delta} \to \tyPrim{real}{\epsilon}
\end{displaymath}

@@

This paper makes the following specific contributions:
\begin{itemize}
\item 
We present a collection of compelling examples of algebraically
indexed types, including a novel type system for geometry, a type
system for tracking taint marking (or provenance) of strings, a
refined type system for information flow based on logic, and a simple
type system for basic uniform continuity analysis.
\item 
We formulate a type system that can either be used as a programming
language in its own right, or as the target of type-based
analyses. The type system consists of the usual type constructors
together with a collection of indexed primitive types, universal and
existential quantification over the indices, and a multi-sorted
equational theory for indices.
\item
We describe a relational semantics for the type system and prove an
analogue of Reynolds' Abstraction Theorem. We give precise conditions
on sets of relation environments under which our Abstraction Theorem
holds.
\item
For each of our main examples we deduce free theorems that are
consequences of our Abstraction Theorem, prove non-definability
results, and derive interesting type isomorphisms.
\end{itemize}
We have fully formalised our framework and Abstraction Theorem in Coq,
using strongly-typed term representations
throughout~\cite{TypedSyntax}.

>>>>

Transformations typically \emph{compose} in various
ways, and the compositions satisfy algebraic laws. For example, 
we can assign a function that computes the area of a circle given its
radius the following polymorphic type:
\[
\mathrm{areaCircle} : \forall\alpha\mathord:\SynGL{1}.\tyPrim{real}{\alpha}\to
\tyPrim{real}{\alpha\cdot\alpha}
\]
This captures the fact that the area of a circle varies as the square
of its radius, i.e., $\mathrm{areaCircle}\;(s\cdot r) =
s^2\cdot \mathrm{areaCircle}\;r$ for any $s\neq 0$.  Here, $\alpha$ is interpreted as
the \emph{units of measure} of the argument to $\mathrm{areaCircle}$,
and `$\cdot$' composes units using the product. We can also add an
inverse operation and identity unit of measure $1$, and then impose the
algebraic laws of abelian groups. This permits identification of, for
example, $\tyPrim{real}{\alpha\cdot\alpha^{-1}}$ with the type
$\tyPrim{real}{1}$ of dimensionless constants.

\paragraph{Abstraction}
In his original paper on parametricity, Reynolds asserted that
\emph{type structure is a syntactic discipline for enforcing levels of
  abstraction}.  We see something analogous here: if all primitive
operations are given types that reflect their behaviour under
translation, then there is no way to `break' this property. For
example, there is no way that $\mathrm{areaTri}$ can depend on the
actual coordinates of its inputs. Furthermore, the distinction between
points and vectors that is often enforced through abstract data
types~\cite{CGAL} is captured here by indices instead. For example,
the operation that takes two points and computes their vector
difference can be assigned the type
$\forall\alpha\mathord:\SynTransl.\tyPrim{vec}\alpha\times\tyPrim{vec}\alpha\to\tyPrim{vec}0$,
reflecting the invariance of the result (a pure vector) under
translations of the point arguments. As a result through types alone
we can, in essence, derive so-called \emph{coordinate-free}
geometry~\cite{CFGADT}.

@@

\paragraph{Relational parametricity}
To derive such invariance and abstraction properties of types, we
adopt the techniques of relational parametricity. Over an underlying
index-erasure semantics we construct binary relations parameterised by
an environment $\rho$ that describes how values of primitive type are
related according to their indices.  For example, values $v$ and $w$
of type $\tyPrim{real}\alpha$ are related when $v$ ``scales to'' $w$
according to an interpretation of $\alpha$ (i.e., $w=\rho(\alpha)\cdot
v$).  Values of polymorphic type are related exactly when they are
related for all possible interpretations of the quantified
variable. For example, values $v$ and $w$ of type
$\forall\alpha\mathord:\SynTransl{2}.\tyPrim{vec}\alpha\to
\tyPrim{vec}\alpha$ are related when they are related at type
$\tyPrim{vec}\alpha\to\tyPrim{vec}\alpha$ for all translations
$t\in\Transl{2}$ associated with $\alpha$.

@@

As it happens, the above examples can all be expressed using unary
functions rather than binary relations: we can write functions that
scale, or translate, one value to another according to their type,
even for higher-order function types. Moreover, the interpretation of
indices can be built up compositionally, by recursion over the
syntactic structure. For example, we have $\rho(\alpha\cdot\beta) =
\rho(\alpha)\cdot\rho(\beta)$ where product of units-of-measure is
interpreted by product of scale factors. For more sophisticated
applications, however, we need the fully generality of binary
relations. As is the case with type polymorphism, it
provides the appropriate meaning for nested quantifiers in types such
as
$(\forall\alpha\mathord:\SynGL{1}.\tyPrim{real}\alpha\to\tyPrim{real}\alpha)\to\tyBool$.
More interestingly, particular applications sometimes require
relations that are not built compositionally---we can, for instance,
use such relations to prove that there are no non-trivial programs of
type
$\forall\alpha\mathord:\SynGL{1}.\tyPrim{real}{\alpha\cdot\alpha}\to\tyPrim{real}\alpha$.
Finally, there are interesting applications in which even the basic
relations are not functional. In a type system in which the index in
$\tyPrim{real}\alpha$ is interpreted not as a unit of measure, but as
a measure of \emph{closeness}, two values $x$ and $y$ of this type are
related if $|x-y| < \rho(\alpha)$ for a positive real number
$\rho(\alpha)$.  Rather beautifully, the standard notion of uniform
continuity can then be expressed as a type:
\begin{displaymath}
  \forall \epsilon \mathord: \mathsf{R}^{>0}.\ \exists \delta\mathord: \mathsf{R}^{>0}.\ \tyPrim{real}{\delta} \to \tyPrim{real}{\epsilon}
\end{displaymath}

@@

This paper makes the following specific contributions:
\begin{itemize}
\item 
We present a collection of compelling examples of algebraically
indexed types, including a novel type system for geometry, a type
system for tracking taint marking (or provenance) of strings, a
refined type system for information flow based on logic, and a simple
type system for basic uniform continuity analysis.
\item 
We formulate a type system that can either be used as a programming
language in its own right, or as the target of type-based
analyses. The type system consists of the usual type constructors
together with a collection of indexed primitive types, universal and
existential quantification over the indices, and a multi-sorted
equational theory for indices.
\item
We describe a relational semantics for the type system and prove an
analogue of Reynolds' Abstraction Theorem. We give precise conditions
on sets of relation environments under which our abstraction theorem
holds.
\item
For each of our main examples we deduce free theorems that are
consequences of our abstraction theorem, prove non-definability
results, and derive interesting type isomorphisms.
\end{itemize}
We have fully formalised our framework and Abstraction Theorem in Coq,
using strongly-typed term representations
throughout~\cite{TypedSyntax}, available from
\\\url{github.com/bobatkey/algebraically-indexed-types/tree/master/coq}
